---
title: "Analysing Clinical Survival Data with OptSurvCutR"
subtitle: "Discovering Biomarker Thresholds and Tier Systems in Liver Disease"
author: "Payton Yau"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Analysing Clinical Survival Data with OptSurvCutR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = TRUE
)

# CRAN-Safe CLI formatting
options(cli.num_colors = 1)
```

```{=html}
<style>
/* Explicitly restrict blockquote styling to document text bodies to shield gridtext */
  .main-container blockquote, .body-content blockquote, article blockquote {
    padding: 15px 20px;
    margin: 0 0 20px;
    font-size: 1.05em;
    border-left: 5px solid #E74C3C; 
    background-color: #fdfdfd;
    color: #444;
    border-radius: 0px 8px 8px 0px;
    box-shadow: 2px 2px 8px rgba(0,0,0,0.05);
  }
  .main-container blockquote p, .body-content blockquote p, article blockquote p {
    margin-bottom: 0;
  }
</style>
```

# Introduction

In clinical research, continuous biomarkers are often simplified into binary groups using arbitrary thresholds like median splits. While convenient, this approach obscures meaningful biological structure. 

The **OptSurvCutR** package is built specifically to discover true, mathematically optimal thresholds. Its core powerhouse is a **Genetic Algorithm**—an evolutionary search engine capable of navigating massive, multi-dimensional data to find 1, 2, or even 4 optimal cut-points simultaneously, a task where traditional methods would freeze.

Furthermore, the package features built-in statistical safety nets, including a 2-Tier Schoenfeld diagnostic check and a comprehensive 4-Tier Validation Stability system, ensuring the thresholds you discover are mathematically defensible.

In this vignette, we apply this workflow to the classic **Mayo Clinic Primary Biliary Cholangitis (PBC)** dataset to evaluate raw **serum bilirubin**, a well-established prognostic marker of liver dysfunction.

---

# 1. Setup & Data Preparation
```{r load-libraries, message=FALSE, warning=FALSE}
library(survival)
library(dplyr)
library(ggplot2)
library(patchwork)
library(knitr)
library(cli)
library(OptSurvCutR)
```

We load the PBC dataset and extract the complete cases. 
```{r load-data}
data(pbc, package = "survival")

# Extract complete cases for our variables of interest
pbc_clean <- na.omit(pbc[, c("time", "status", "bili")])

# Define the survival event (2 = dead, everything else is censored)
pbc_clean$event <- ifelse(pbc_clean$status == 2, 1, 0)

head(pbc_clean)
```

> **💡 Key Insight: Dealing with Skewed Data**
> Raw clinical lab values are often heavily right-skewed (our `bili` data ranges from 0.3 to 28.0, with most patients bunched at the bottom). Because `OptSurvCutR` evaluates survival using the Log-rank test—which looks at the *rank order* of patients rather than their absolute numerical distance—the algorithm is mathematically immune to this skew. You do not need to manually log-transform your data; just feed the raw clinical values directly into the engine!
> 
> *Advanced Note:* If your specific research workflow requires mathematically smoothing the risk curves prior to cut-point discovery, commonly used R packages for data smoothing include **`mgcv`** (for Generalized Additive Models / GAMs) and **`splines`**.

---

# 2. The Three-Step Analysis Workflow

## Step 1: Determine the Optimal Number of Cuts

Before we ask *where* the cut-points are, we need to ask the math *how many* cut-points actually exist in our data. 

### Choosing Your Information Criterion
The algorithm evaluates models using an Information Criterion, which balances accuracy against complexity:

| Criterion | What it does | Best Use Case |
| :--- | :--- | :--- |
| **AIC** | Light penalty for complexity. | Exploratory research; allows more cuts. |
| **AICc** | Corrects AIC for small sample sizes. | Small clinical datasets (N < 200). |
| **BIC** | Heavy penalty for complexity. | Confirmatory research; prefers simpler, highly robust models. |
```{r find-number-BIC}
num_res <- find_cutpoint_number(
  # ==========================================
  # 1. CORE DATA INPUTS
  # ==========================================
  data = pbc_clean, # The patient dataset
  predictor = "bili", # The continuous biomarker
  outcome_time = "time", # Survival time column
  outcome_event = "event", # Survival event column

  # ==========================================
  # 2. SEARCH ENGINE SETTINGS
  # ==========================================
  method = "genetic", # The evolutionary search engine
  criterion = "BIC", # Evaluates model fit (BIC heavily penalizes over-complexity)
  max_cuts = 3, # Test models ranging from 0 to 3 cuts
  nmin = 0.25, # SAFETY LIMIT: Minimum 10% of patients per group
  n_perm = 50,
  # ==========================================
  # 3. THE "R-GROUND" SETTINGS (Advanced Genetic Tuning)
  # ==========================================
  max.generations = 100, # LIFESPAN: How many evolutionary cycles to run
  pop.size = 100, # SEARCH PARTY: Number of random guesses per generation
  boundary.enforcement = 1, # EDGE CONTROL: 1 = Soft boundary (allows safe edge exploration)
  seed = 123 # REPRODUCIBILITY: Locks the random number generator
)

plot(num_res) +
  geom_smooth(method = "loess", se = FALSE, colour = "darkgrey", linetype = "dashed", alpha = 0.5)
```

### ⚠️ Step 1 Warning System: Algorithm Failure
If the genetic algorithm is suffocated by overly strict safety parameters, it will output the following warning:
> `"No valid cut-points found for X cut(s) due to genetic algorithm failure or constraints."`

**How to fix it:** If you see this, lower your `nmin` (e.g., to 0.05), ensure `boundary.enforcement = 1`, or increase `max.generations` to give the algorithm more room to search.

> **⚠️ Important: The Clinical & Statistical Overrule (Why we step down to 2 cuts)**
> Looking at the BIC trend above, the math might suggest that **3 cuts (4 groups)** provides the best raw fit for this specific dataset. However, statistical fit does not always equal biological stability. 
> 
> If we blindly pushed a 3-cut model through our rigorous Bootstrap Validation (Step 3), the resulting groups would be too small. The cut-points would bounce wildly across samples, causing the confidence intervals to overlap and the risk cohorts to bleed into one another (resulting in a **Tier 3 or Tier 4 Unstable** rating). The algorithm would be overfitting to sample noise.
> 
> To prevent this and ensure our model discovers mathematically distinct, well-separated subpopulations, we will intentionally overrule the initial mathematical suggestion. In Step 2, we will drop down to **2 cuts (3 groups)**. This mirrors standard clinical triaging (Normal, Elevated, Severe) and gives the groups enough data density to guarantee a robust, high-tier validation.

---

## Step 2: Pinpointing the Exact Cut-point Values

This step is the heart of `OptSurvCutR`. We ask the search engine to physically locate the optimal dividing lines. 

### Choosing Your Engine
You must select the correct mathematical engine based on your goal:

| Engine | How it works | Best for | Drawback |
| :--- | :--- | :--- | :--- |
| `systematic` | Brute-force checks every single number. | Finding exactly **1 cut-point**. | Mathematically impossible to use for 2+ cuts (takes hours/days). |
| `genetic` | Drops random guesses and "evolves" the best answer. | Finding **2, 3, or 4 cut-points** instantly. | Can occasionally fail if parameters (`nmin`) are too strict. |
```{r find-cutpoint}
cut_res <- find_cutpoint(
  # ==========================================
  # 1. CORE DATA INPUTS
  # ==========================================
  data = pbc_clean,
  predictor = "bili",
  outcome_time = "time",
  outcome_event = "event",

  # ==========================================
  # 2. SEARCH ENGINE SETTINGS
  # ==========================================
  method = "genetic", # Crucial for finding >1 cut-point
  criterion = "logrank", # Maximizes survival difference between groups
  num_cuts = 2, # Forcing 2 cuts (3 groups) for demonstration and stability
  nmin = 0.25, # SAFETY LIMIT: Forces every group to contain >= 25% of patients
  n_perm = 100, # PERMUTATION TEST: Scrambles the data to calculate an honest p-value

  # ==========================================
  # 3. THE "R-GROUND" SETTINGS (Advanced Genetic Tuning)
  # ==========================================
  max.generations = 100, # LIFESPAN: Increase if the algorithm fails to converge
  pop.size = 100, # SEARCH PARTY: Increase to 500+ for massive datasets
  boundary.enforcement = 1, # EDGE CONTROL: Soft boundaries prevent the algorithm from crashing near edges
  seed = 123, # REPRODUCIBILITY
  n_cores = 1 # SPEED: Adjust based on your machine's CPU
)

summary(cut_res)
```

### ⚠️ Step 2 Warning System: The 2-Tier Schoenfeld Diagnostic
When `summary()` is called, the package automatically evaluates your discovered cut-points against the fundamental rule of survival analysis: **The Proportional Hazards Assumption**. It does this using the Schoenfeld residuals test, which checks if the risk between your groups remains parallel over time. It flags the result using a 2-Tier Diagnostic System (`── 3. Time-Dependent Diagnostics (Schoenfeld) ──`):

| Diagnostic Status | What it Means (Schoenfeld Test) | Clinical Implication |
| :--- | :--- | :--- |
| **Tier 1 (Proportional)** | The proportional hazards assumption holds ($p > 0.05$). | Your cut-points remain equally deadly/protective on Day 1 as they do on Day 1,000. Reporting a standard Hazard Ratio is valid. |
| **Tier 2 (Time-Varying)** | The hazard ratio changes significantly over time ($p < 0.05$). | Your cut-point is still biologically valid and separates survival, but its predictive power shifts (e.g., highly lethal early on, but fades later). You may need to report time-dependent Hazard Ratios. |

### Distribution Plot with KDE Interpolation
```{r plot_find-cutpoint}
plot(cut_res, type = "distribution") +
  geom_rug(alpha = 0.5) +
  labs(caption = "KDE interpolation reveals the continuous distribution of bilirubin")
```

> **💡 Key Insight: Reading the Distribution Plot**
> The blue shaded area is the density of your patients. The orange dashed lines are the exact biological thresholds the algorithm discovered. The algorithm doesn't just cut the population evenly; it places the lines exactly where the *risk of death* sharply increases, isolating the healthy baseline from the severe outliers.

---

## Step 3: Validate the Threshold Stability

Finding a cut-point in one dataset is easy; proving that cut-point is real and will hold up in future patients is hard. We use "Bootstrap Resampling" to stress-test our findings.
```{r validate-cutpoint}
val_res <- validate_cutpoint(
  # ==========================================
  # 1. VALIDATION INPUTS
  # ==========================================
  cutpoint_result = cut_res, # The exact result object you generated in Step 2
  num_replicates = 100, # BOOTSTRAP: How many times to scramble and re-test. 500+ is the gold standard
  n_cores = 1, # SPEED: Number of CPU cores to use

  # ==========================================
  # 2. THE "R-GROUND" SETTINGS (Must match Step 2)
  # ==========================================
  max.generations = 100, # Must match or exceed Step 2 to ensure stability during bootstrap
  pop.size = 100, # Must match or exceed Step 2
  boundary.enforcement = 1, # Must match Step 2
  seed = 123 # REPRODUCIBILITY
)

summary(val_res)

knitr::kable(val_res$confidence_intervals, digits = 3, caption = "Stability Intervals for Bilirubin Risk Thresholds")

plot(val_res) +
  labs(caption = "Density interpolation of bootstrap-derived cut-points")
```

### ⚠️ Step 3 Warning System: The 4-Tier Stability Assessment
Bootstrapping is highly aggressive. Once complete, `OptSurvCutR` evaluates both the **Precision** (Maximum CI Width) and the **Validity** (Interval Overlap), automatically categorizing your threshold's stability into one of four diagnostic tiers:

*   **Tier 1 (OPTIMAL):** Narrow confidence intervals (CI Width < 30%) and clean separation. The thresholds are rock-solid, highly reproducible, and distinctly divide the population.
*   **Tier 2 (DISTINCT):** Zero overlap between intervals, regardless of exact CI Width. The exact threshold value may bounce across samples (moderate to high variance), but the resulting biological subpopulations remain fundamentally separate and never bleed into one another. 
*   **Tier 3 (CAUTION):** Moderate variance (CI Width 30%–60%) **WITH** overlapping intervals. The threshold is somewhat sensitive to sample changes, and the resulting risk cohorts may blend together in the 'grey zones'.
*   **Tier 4 (UNSTABLE):** High variance (CI Width > 60%) **WITH** overlapping intervals. The exact numbers are highly volatile, and the groups bleed into each other completely, indicating potential over-fitting to noise.

> **💡 Key Insight: Why do we Bootstrap?**
> Imagine we randomly re-sampled your patient cohort (with replacement) to create 500 slightly different "alternate reality" datasets. Would your cut-points completely change or hold steady? Bootstrapping simulates this process. The tight, peaked density curves in the plot above prove that our thresholds are a genuine biological signal, not just a mathematical fluke of this specific patient sample!

---

# 3. Advanced Visualisation & Reporting

## 3.1 Group Composition Table
```{r group-composition}
final_dataset <- plot(cut_res, return_data = TRUE)

composition_table <- final_dataset %>%
  group_by(group) %>%
  summarise(
    Bilirubin_Range = paste0(
      round(min(factor), 3), " – ", round(max(factor), 3)
    ),
    N = n(),
    Deaths = sum(event)
  )

knitr::kable(composition_table, caption = "Composition of Discovered Bilirubin Risk Groups")
```

## 3.2 Hazard Ratio Forest Plot (`type = "forest"`)
```{r plot-forest, fig.width=7, fig.height=3.5}
plot(cut_res,
  type = "forest",
  main = "Mortality Hazard Ratios Relative to Group 1 (Baseline)"
)
```

> **⚠️ Important: The Hazard Ratio**
> The forest plot explicitly quantifies the increased risk of mortality. Group 3 (the severe impairment group) shows a massive, statistically significant hazard ratio compared to the baseline group, confirming the clinical urgency of crossing this biological threshold.

## 3.3 Time-Dependent Diagnostics (`type = "diagnostic"`)
```{r plot-diagnostic}
plot(cut_res, type = "diagnostic")
```

## 3.4 2D Cut-point Stability Surface (plot_validation())
Rather than relying on blocky grids, this specialized diagnostic plot renders a smooth contour elevation surface. It visualizes exactly how your discovered cut-points cluster across bootsrapped replicates or validation folds.

```{r plot_validation}
# Evaluates the stability cluster between your second and third clinical thresholds
plot_validation(val_res,
  focus_cuts = c(1, 2),
  main = "Upper Tier Threshold Stability Map"
)
```

## 3.6 Standalone Kaplan-Meier Survival Curves (`type = "outcome"`)
```{r plot-km, fig.width=8, fig.height=6}
plot(cut_res,
  type = "outcome",
  title = "PBC Survival by Bilirubin Tier",
  xlab = "Follow-up Time (Days)",
  ylab = "Overall Survival Probability",
  legend.title = "Bilirubin Tier"
)
```

> **💡 Key Insight: The Kaplan-Meier Step Function**
> This is the most famous plot in oncology and hepatology. Each time a patient experiences the event (death), the line steps down. The visual "gap" between the red, green, and blue lines represents the life-saving or life-threatening impact of being in a specific bilirubin tier. Wide gaps that don't cross over each other visually confirm that the algorithm successfully found distinctly different biological risk groups.

## 3.7 The Escape Hatch: Exporting Your Stratified Data

While `OptSurvCutR` provides beautiful plots, you will likely need to export your newly stratified patients to run custom models or share with collaborators. 

By passing `return_data = TRUE` to the plot function, you can extract your exact original dataset, but with a new `group` column added that assigns every patient to their mathematically discovered tier!
```{r export-data}
# Extract the classified dataset
stratified_patients <- plot(cut_res, return_data = TRUE)

# View the new data
head(stratified_patients[, c("time", "event", "factor", "group")])

# Save to Excel/CSV for your team!
# write.csv(stratified_patients, "PBC_Patients_Tiered.csv", row.names = FALSE)
```

---

# 4. Conclusion

This vignette demonstrates how **OptSurvCutR** bridges the gap between raw statistical optimization and clinical reality. By pairing advanced genetic algorithms with comprehensive diagnostic warnings, researchers can confidently discover robust, stable risk thresholds directly from un-transformed clinical laboratory data.

- **Install the package:** `remotes::install_github("paytonyau/OptSurvCutR")`
- **Report issues or suggest features** on our [GitHub page](https://github.com/paytonyau/OptSurvCutR).

---

# 5. Session Information
```{r session_info}
sessionInfo()
```
