---
title: "Analysing Clinical Survival Data with OptSurvCutR"
subtitle: "Discovering Biomarker Thresholds and Tier Systems in Liver Disease"
author: "Payton Yau"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Analysing Clinical Survival Data with OptSurvCutR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4.5,
  fig.align = "center",
  warning = TRUE,
  message = TRUE
)

# CRAN-Safe CLI formatting
options(cli.num_colors = 1)
```

```{=html}
<style>
/* Explicitly restrict blockquote styling to document text bodies to shield gridtext */
  .main-container blockquote, .body-content blockquote, article blockquote {
    padding: 15px 20px;
    margin: 0 0 20px;
    font-size: 1.05em;
    border-left: 5px solid #E74C3C; 
    background-color: #fdfdfd;
    color: #444;
    border-radius: 0px 8px 8px 0px;
    box-shadow: 2px 2px 8px rgba(0,0,0,0.05);
  }
  .main-container blockquote p, .body-content blockquote p, article blockquote p {
    margin-bottom: 0;
  }
</style>
```

# Introduction

In clinical research, continuous biomarkers are often simplified into binary groups using arbitrary thresholds like median splits. While convenient, this approach obscures meaningful biological structure.

The **OptSurvCutR** package is built specifically to discover true, mathematically optimal thresholds. Its core powerhouse is a **Genetic Algorithm**—an evolutionary search engine capable of navigating massive, multi-dimensional data to find optimal cut-points simultaneously, a task where traditional methods would freeze.

Furthermore, the package features built-in statistical safety nets, including a 2-Tier Schoenfeld diagnostic check and a comprehensive 4-Tier Validation Stability system. Crucially, the engine allows for **covariate adjustment**, ensuring that the discovered biomarker thresholds provide independent prognostic value even when controlling for established clinical confounders like age or disease staging.

In this vignette, we apply this workflow to the classic **Mayo Clinic Primary Biliary Cholangitis (PBC)** dataset to evaluate raw **serum bilirubin**, controlling for age and edema.

------------------------------------------------------------------------

# 1. Setup & Data Preparation

To begin the analysis, we must first load our workspace environment with the required data-wrangling libraries, graphics systems, and the underlying survival and optimisation packages.

```{r load-libraries, message=FALSE, warning=FALSE}
library(survival)
library(dplyr)
library(ggplot2)
library(patchwork)
library(knitr)
library(cli)
library(OptSurvCutR)
```

Next, we pull the historical baseline dataset into memory, strip out clinical records containing missing entries across our features of interest, and map all active events safely to isolate true endpoints.

```{r load-data}
data(pbc, package = "survival")

# Extract complete cases for our predictor, outcome, and clinical covariates
pbc_clean <- na.omit(pbc[, c("time", "status", "bili", "age", "sex", "edema")])

# Define the all-cause mortality survival event (1 or 2 = endpoint, 0 = censored)
pbc_clean$event <- ifelse(pbc_clean$status %in% c(1, 2), 1, 0)

head(pbc_clean)
```

> **💡 Key Insight: Dealing with Skewed Data & Confounders** Raw clinical lab values are often heavily right-skewed. Because `OptSurvCutR` evaluates survival using the Log-rank/Cox rank infrastructure—which looks at the *rank order* of patients rather than absolute numerical distance—the algorithm is mathematically immune to this skew.
>
> By passing `covariates = c("age","sex", "edema")`, the optimisation routine transitions from a univariate survival split to an adjusted Cox proportional hazards model framework, ensuring your cut-points reflect true baseline risk alterations independent of baseline age or physical edema presentation.

------------------------------------------------------------------------

# 2. The Three-Step Analysis Workflow

## Step 1: Determine the Optimal Number of Cuts

Before we ask *where* the cut-points are, we need to ask the math *how many* cut-points actually exist in our data when adjusting for our baseline covariates.

### Choosing Your Information Criterion

The algorithm evaluates models using an Information Criterion, which balances accuracy against complexity:

| Criterion | What it does | Best Use Case |
|:-----------------------|:-----------------------|:-----------------------|
| **AIC** | Light penalty for complexity. | Exploratory research; allows more cuts. |
| **AICc** | Corrects AIC for small sample sizes. | Small clinical datasets (N \< 200). |
| **BIC** | Heavy penalty for complexity. | Confirmatory research; prefers simpler, highly robust models. |

We deploy the multi-model search check to calculate fitness scores across various grouping parameters, enforcing safety boundaries so that cohorts don't become unsustainably small.

```{r find-number-BIC}
num_res <- find_cutpoint_number(
  # ==========================================
  # 1. CORE DATA & COVARIATE INPUTS
  # ==========================================
  data = pbc_clean,           # The patient dataset
  predictor = "bili",         # The continuous biomarker
  outcome_time = "time",      # Survival time column
  outcome_event = "event",    # Survival event column
  covariates = c("age","sex","edema"), # Confounders to control for

  # ==========================================
  # 2. SEARCH ENGINE SETTINGS
  # ==========================================
  method = "genetic",         # The evolutionary search engine
  criterion = "AIC",          # Evaluates adjusted model fit
  max_cuts = 5,               # Test models ranging from 0 to 5 cuts
  nmin = 0.15,                # SAFETY LIMIT: Minimum 15% of patients per group
   
  # ==========================================
  # 3. ADVANCED GENETIC TUNING
  # ==========================================
  max.generations = NULL,     # LIFESPAN: How many evolutionary cycles to run
  pop.size = NULL,            # SEARCH PARTY: Number of random guesses per generation
  boundary.enforcement = 2,   # EDGE CONTROL: 2 = Soft boundary configuration
  seed = 123                  # REPRODUCIBILITY
)

```

Here, we use `summary()` for detailed information.

```{r find-number-BIC-summary}
summary(num_res)
```

To easily track the inflection points where adding more thresholds yields diminishing mathematical returns, we plot the complete multi-model fitness curve.

```{r loess}
plot(num_res) +
  geom_smooth(method = "loess", 
              se = FALSE, 
              colour = "darkgrey", 
              linetype = "dashed", 
              alpha = 0.5)
```

**Interpretation:** The model minimising the information landscape sits clearly at **2 cut-points** (which divides the patient cohort into 3 distinct risk groups). This confirms that even after stripping away the confounding baseline effects of age and sex, a multi-tier group model is statistically superior to a traditional median binary split.

## Step 2: Pinpointing the Exact Cut-point Values

This step is the heart of `OptSurvCutR`. We ask the search engine to physically locate the optimal dividing lines while holding our adjusted covariates constant.

We initiate the core optimisation search loop, directing the engine to settle on an integer space topology that maximises group risk divergence under active covariate adjustment.

```{r find-cutpoint}
cut_res <- find_cutpoint(
  # ==========================================
  # 1. CORE DATA & COVARIATE INPUTS
  # ==========================================
  data = pbc_clean,
  predictor = "bili",
  outcome_time = "time",
  outcome_event = "event",
  covariates = c("age","sex","edema"), # Adjusted during location discovery

  # ==========================================
  # 2. SEARCH ENGINE SETTINGS
  # ==========================================
  method = "genetic", 
  criterion = "logrank", 
  num_cuts = 3,       # Setting 3 cuts for clinical tier separation
  nmin = 0.15,         # SAFETY LIMIT: Minimum patients per risk cohort
  n_perm = 20,        # Streamlined permutation cycle

  # ==========================================
  # 3. ADVANCED GENETIC TUNING
  # ==========================================
  max.generations = NULL,   
  pop.size = NULL,          
  boundary.enforcement = 2, 
  seed = 123,               
  n_cores = 1               
)
```

Now, we call the optimisation summary report to extract hazard thresholds, coefficients, and preliminary model diagnostics.

```{r find-cutpoint-summary}
summary(cut_res)
```

### 🔍 Automated Diagnostic Check: The 2-Tier Schoenfeld Diagnostic

When evaluating `summary()`, the package automatically processes a Schoenfeld residuals validation check against your adjusted model matrix, verifying if the Proportional Hazards assumption holds:

| Diagnostic Status | What it Means (Schoenfeld Test) | Clinical Implication |
|:-----------------------|:-----------------------|:-----------------------|
| **Tier 1 (Proportional)** | The proportional hazards assumption holds ($p > 0.05$). | Your cut-points remain equally deadly/protective on Day 1 as they do on Day 1,000. Reporting a standard adjusted Hazard Ratio is valid. |
| **Tier 2 (Time-Varying)** | The hazard ratio changes significantly over time ($p < 0.05$). | Your cut-point is still biologically valid, but its predictive power shifts over time. You may need to report time-dependent Hazard Ratios. |

To visualise where these calculated cut-points slice across the patient density profile, we map our continuous marker distribution.

```{r plot_find-cutpoint}
plot(cut_res, type = "distribution") +
  geom_rug(alpha = 0.5) +
  labs(caption = "KDE interpolation reveals the covariate-adjusted thresholds of bilirubin")
```

------------------------------------------------------------------------

## Step 3: Validate the Threshold Stability

We use Bootstrap Resampling to ensure our covariate-adjusted thresholds are real and completely independent of sample noise.

We run repetitive bootstrap sampling loops, forcing the optimisation engine to re-verify the marker thresholds across randomised slices of our cohort.

```{r validate-cutpoint}
val_res <- validate_cutpoint(
  # ==========================================
  # 1. VALIDATION INPUTS
  # ==========================================
  cutpoint_result = cut_res, 
  num_replicates = 30,       # 500+ is the standard for final publication
  n_cores = 1, 

  # ==========================================
  # 2. ADVANCED SETTINGS (Must match Step 2)
  # ==========================================
  max.generations = NULL, 
  pop.size = NULL, 
  boundary.enforcement = 2, 
  seed = 123 
)
```

We look at the aggregated validation metrics to review the underlying variance metrics across our calculated thresholds.

```{r validate-cutpoint-summary}
summary(val_res)
```

### 🔍 Automated Performance Grading: The 4-Tier Stability Assessment

Following bootstrap calculations, `OptSurvCutR` evaluates both your **Precision** (Confidence Interval width relative to overall data range) and **Validity** (absence of interval overlap), grading model health into four distinct tiers:

1.  **Tier 1 (OPTIMAL):** Narrow intervals (CI Width \< 30%) paired with complete cohort separation.
2.  **Tier 2 (DISTINCT):** Zero overlap between confidence zones regardless of width. Thresholds may float slightly, but risk boundaries remain clear.
3.  **Tier 3 (CAUTION):** Moderate parameter variance (CI Width 30%–60%) accompanied by overlapping intervals.
4.  **Tier 4 (UNSTABLE):** High variance (CI Width \> 60%) with overlapping intervals, signaling severe data over-fitting.

For precise documentation, we compile our numerical confidence intervals directly into a clean markdown summary table.

```{r validate-cutpoint-knitr}
knitr::kable(val_res$confidence_intervals, digits = 3, caption = "Covariate-Adjusted Stability Intervals for Bilirubin Risk Thresholds")
```

To see if our thresholds remain tightly anchored or scatter across the iterations, we inspect the bootstrap density landscape.

```{r validate-cutpoint-plot}
plot(val_res) +
  labs(caption = "Density interpolation of bootstrap-derived cut-points under adjustment")
```

------------------------------------------------------------------------

# 3. Advanced Visualisation & Reporting

## 3.1 Group Composition Table

We can check the physical distributions and average baseline parameters across our newly established medical risk tiers.

```{r group-composition}
final_dataset <- plot(cut_res, return_data = TRUE)

composition_table <- final_dataset %>%
  group_by(group) %>%
  summarise(
    Bilirubin_Range = paste0(
      round(min(factor), 3), " – ", round(max(factor), 3)
    ),
    N = n(),
    Deaths = sum(event),
    Mean_Age = round(mean(age), 1),
    Edema_Rate = round(mean(edema), 2)
  )

knitr::kable(composition_table, caption = "Composition and Covariate Profiles of Discovered Bilirubin Risk Groups")
```

## 3.2 Adjusted Hazard Ratio Forest Plot (`type = "forest"`)

We can easily plot a custom forest chart to display our adjusted hazard models relative to the low-risk baseline tier.

```{r plot-forest, fig.width=7, fig.height=3.5}
plot(cut_res,
  type = "forest",
  main = "Adjusted Mortality Hazard Ratios Relative to Group 1"
)
```

## 3.3 Time-Dependent Diagnostics (`type = "diagnostic"`)

We track whether our optimisation groups safely adhere to proportional variance targets across the complete time scope.

```{r plot-diagnostic}
plot(cut_res, type = "diagnostic")
```

## 3.4 2D Cut-point Stability Surface (plot_validation())

Rather than relying on blocky grids, `plot_validation()` allows us to project high-dimensional bootstrap convergence horizons onto a smooth 2D continuous contour map, highlighting exactly where our optimal threshold combinations cluster.

```{r plot_validation}
plot_validation(val_res,
  focus_cuts = c(1, 2),
  main = "Upper Tier Threshold Stability Map"
)
```

## 3.5 Adjusted Kaplan-Meier Survival Curves (`type = "outcome"`)

We evaluate the clear divergence in survival over time by mapping the adjusted step functions for our groups.

```{r plot-km}
plot(cut_res,
  type = "outcome",
  title = "PBC Survival by Bilirubin Tier (Covariate Adjusted)",
  xlab = "Follow-up Time (Days)",
  ylab = "Overall Survival Probability",
  legend.title = "Bilirubin Tier"
)
```

## 3.6 Exporting Your Stratified Data

By passing `return_data = TRUE`, you can seamlessly extract your original clinical dataset paired with the newly calculated `group` and original covariate column assignments.

```{r export-data}
stratified_patients <- plot(cut_res, return_data = TRUE)
head(stratified_patients[, c("time", "event", "factor", "age", "sex","edema", "group")])
```

------------------------------------------------------------------------

# 4. Conclusion

This vignette demonstrates how **OptSurvCutR** controls for clinical confounding factors during optimal threshold selection. By providing a streamlined pipeline for covariate-adjusted optimisation, clinical scientists can establish reliable diagnostic thresholds that retain absolute structural validity across independent validation cohorts.

------------------------------------------------------------------------

# 5. Session Information

Finally, we print our system specifications, environmental packages, and runtime tracking indices to ensure perfect documentation reproducibility.

```{r session_info}
sessionInfo()
```

\`\`\`
