---
title: "Stratifying Colorectal Cancer Patients by Lymph Node Involvement"
subtitle: "Covariate-Adjusted Survival Analysis with OptSurvCutR"
author: "Payton Yau"
date: "2026-06-08"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Stratifying Colorectal Cancer Patients by Lymph Node Involvement}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
# Set global options for all code chunks in the document.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = TRUE
)
```

### About the Package

The **`OptSurvCutR`** package provides a comprehensive, data-driven
workflow for survival analysis, specifically designed for scenarios
where a single cut-point is not sufficient. Its core functionality is to
discover and validate one or more optimal cut-points for continuous
variables in time-to-event data.

A primary asset of `OptSurvCutR` is its ability to perform threshold
discovery **while adjusting for critical covariates** (e.g., patient
age, sex, or treatment arms). This ensures that the discovered risk
groups carry independent prognostic value, preventing confounding
clinical factors from muddying your biomarker’s true signal.

------------------------------------------------------------------------

# Introduction

This tutorial demonstrates a complete, end-to-end workflow for finding
independent, covariate-adjusted survival thresholds using a built-in
clinical registry dataset.

### The Scientific Background

Metastatic spread to regional lymph nodes represents one of the most
critical staging transformations and prognostic determinants in colon
adenocarcinoma. In this workflow, we will analyze standard clinical
registry profiles to determine if the continuous count of positive lymph
nodes can independently predict time-to-recurrence or overall survival.

The absolute count of positive lymph nodes introduces an interesting
clinical stratification problem:

1.  **Low Burden Tier:** Minimal to zero nodal involvement typically
    signals standard localized disease with high curative resection
    potential.

2.  **High Burden Tier:** Past a specific clinical inflection point,
    nodal presence transitions from localized regional clusters to
    aggressive systemic failure markers.

Using data from a landmark clinical trial tracking adjuvant therapies
for stage III colon cancer, this vignette handles a robust, multivariate
analysis to isolate data-driven thresholds for positive node counts
while rigorously adjusting for patient age, sex, and tumor
differentiation grade.

### Analysis Workflow

This guide covers a clean, 6-step pipeline:

1.  **Setup and Configuration**: Preparing the environment and setting
    up variables.
2.  **Data Curation**: Loading built-in assets and enforcing a 5-year
    (60-month) timeline cap.
3.  **Number Selection**: Finding the statistically optimal number of
    cuts under covariate pressure.
4.  **Threshold Discovery**: Pinpointing exact cut-point locations via a
    Systematic 1D Grid.
5.  **Stability Assessment**: Validating the discovered boundaries with
    a 4-Tier bootstrap engine.
6.  **Visualization & Interpretation**: Generating publication-ready
    survival curves, adjusted forest plots, and continuous 1D trajectory
    maps.

------------------------------------------------------------------------

## 1. Setup and Configuration

### Loading Libraries

We begin by loading the explicit suite of tools needed to execute the
modeling and manage data piping.

```{r load-libraries}
library(dplyr)
library(survival)
library(survminer)
library(ggplot2)
library(patchwork)
library(knitr)
library(cli)
library(OptSurvCutR)

# Enforce uniform CRAN-safe text formatting logs
options(cli.num_colors = 1)
```

### Defining Target Parameters

We declare our clinical predictor target and confounding adjusters
upfront to maintain clean, reusable variables throughout downstream code
chunks.

```{r configuration}
# Continuous predictor column name (Number of positive lymph nodes)
predictor_name_to_analyze <- "nodes"

# Confounding clinical covariates to adjust for inside the Cox engine
# differ: differentiation grade (1=well, 2=moderate, 3=poor)
covariates_to_adjust_for <- c("age", "sex", "differ")
```

------------------------------------------------------------------------

## 2. Data Loading and Preparation

We load the `colon` dataset directly from the `survival` package. This
data frame tracks chemotherapy adjuvants in stage III colon cancer.
Because the raw dataset records multiple entries per patient mapping
distinct endpoints (recurrence vs. death), we filter strictly for cancer
recurrence records (`etype == 1`). We will then apply a 5-year
administrative censoring cap and filter for complete cases across all
targeted variables.

```{r load-data}
data("colon", package = "survival")

# --- Process and Clean the Case Cohort ---
analysis_data <- colon %>%
  filter(etype == 1) %>% # Isolate recurrence records strictly
  select(
    patient_id = id,
    time_days = time,
    status,
    nodes = all_of(predictor_name_to_analyze),
    all_of(covariates_to_adjust_for)
  ) %>%
  mutate(
    # 1. Convert baseline days timeline into months for uniform layout
    time_months = time_days / 30.4375,

    # 2. Apply 5-year censoring: mark events beyond 60 months as censored (0)
    status_final = ifelse(time_months > 60, 0, status),

    # 3. Cap follow-up times cleanly at the 60-month milestone
    time_final = pmin(time_months, 60)
  ) %>%
  select(
    patient_id,
    time = time_final,
    status = status_final,
    nodes,
    all_of(covariates_to_adjust_for)
  ) %>%
  filter(complete.cases(time, status, nodes, across(all_of(covariates_to_adjust_for))))

# Preview the analysis-ready matrix
head(analysis_data)
```

> **💡 Methodological Insight: Should High-Throughput Biomarker Data be
> Smoothed?** Discrete count or high-throughput omics datasets are
> frequently sparse, zero-inflated, and heavily skewed. **It is highly
> recommended not to smooth your predictor vectors** (e.g., using
> splines or data blurring steps) prior to running `OptSurvCutR`.
>
> Data smoothing can artificially mask and destroy the sharp biological
> boundaries your model seeks to uncover. Instead, pass standard raw
> expressions directly into the package. The underlying optimization
> framework is natively engineered to traverse erratic data surfaces.

------------------------------------------------------------------------

## 3. Covariate-Adjusted Cut-point Discovery

### Step 3.1: Select the Statistically Justified Number of Cut-points

We first pass our data to `find_cutpoint_number()`. This tool loops
through potential architectures (0 to 5 cut-points) while adjusting for
age, sex, and differentiation grade at every iteration. It uses the
Bayesian Information Criterion (BIC) to heavily penalize
over-parameterization. By setting `pop.size = NULL` and
`max.generations = NULL`, the package deploys its newly engineered
dynamic auto-scaling optimization engine.

```{r find-number-Adjusted-Analysis}
number_result_bic <- find_cutpoint_number(
  # ==========================================
  # 1. CORE DATA INPUTS
  # ==========================================
  data = analysis_data,
  predictor = "nodes",
  outcome_time = "time",
  outcome_event = "status",
  covariates = covariates_to_adjust_for,
  # ==========================================
  # 2. SEARCH ENGINE SETTINGS
  # ==========================================
  method = "genetic", # The evolutionary search engine
  criterion = "BIC", # Evaluates model fit (BIC heavily penalizes over-complexity)
  max_cuts = 5, # Test models ranging from 0 to 7 cuts
  nmin = 0.1, # SAFETY LIMIT: Target minimum 10% of patients per group
  # ==========================================
  # 3. AUTOMATED DYNAMIC AUTO-SCALING TUNING
  # ==========================================
  max.generations = NULL, 
  pop.size = NULL, 
  boundary.enforcement = 2, # EDGE CONTROL: Hard restrictions lock optimization bounds
  seed = 123 # REPRODUCIBILITY: Locks the random number generator
)

summary(number_result_bic)
```

```{r plot-number-adj, fig.cap="Bayesian Information Criterion (BIC) landscape across cut architectures. The lowest point denotes the ideal group count."}
# Visualize the model complexity penalty profile
plot(number_result_bic)
```

> **Methodological Note on Missing Model Levels (Cuts 4–5):** \> When
> examining the model selection layout, you will note that candidate
> loops for cuts 4, 5, 6, and 7 are missing from the table. This is an
> intentional security architecture within `OptSurvCutR`. Based on an
> input `nmin = 0.1` over our filtered pool of 888 complete cases, the
> absolute minimum cell floor is set to **88 subjects**
> ($\lfloor 0.1 \times 888 \rfloor = 88$).
>
> High-dimensional cuts require a substantial pool of unique variable
> partitions ($N \ge (k+1) \times \text{nmin}$). Because lymph node
> counts (`nodes`) are highly zero-inflated integers, the discrete
> patient clumps leave insufficient coordinate variety in the long right
> tail to support 5 or more distinct sub-cohorts without violating the
> 88-patient floor. Rather than throwing a fatal error or fitting
> unstable Cox matrices with empty cells, the package throws a clean UX
> alert message and drops the impossible configurations to deliver a
> reliable, parsimonious result. The minimum of the valid information
> landscape sits clearly at **2 cut-point**.

### Step 3.2: Pinpoint Adjusted Cut-point Coordinates (2-Cut Systematic Surface)

With 2 cut established as our mathematical target, we invoke
`find_cutpoint()` using the **systematic method** using the systematic
method to execute an exhaustive grid search and identify the joint
optimal boundaries.

```{r find-cuts-Adjusted-Analysis}
optimal_n_cuts <- number_result_bic$optimal_num_cuts

cutpoint_result_adj <- find_cutpoint(
  # ==========================================
  # 1. CORE DATA INPUTS
  # ==========================================
  data = analysis_data,
  predictor = "nodes",
  outcome_time = "time",
  outcome_event = "status",
  covariates = covariates_to_adjust_for,
  num_cuts = optimal_n_cuts,
  # ==========================================
  # 2. SEARCH ENGINE SETTINGS
  # ==========================================
  method = "systematic", # Standard warning-free systematic search for 1 cut
  criterion = "logrank",
  nmin = 0.1,
  n_perm = 100, # Scaled for swift vignette building
  # ==========================================
  # 3. TUNING OVERRIDES
  # ==========================================
  seed = 123, # REPRODUCIBILITY
  n_cores = 1 # SPEED: Adjust based on your machine's CPU
)

summary(cutpoint_result_adj)
```

### 🔍 Automated Diagnostic Check: The 2-Tier Schoenfeld Diagnostic

When evaluating `summary()`, the package automatically processes a
\***Schoenfeld residuals validation** check against your adjusted model
matrix, verifying if the Proportional Hazards assumption holds:

- **Tier 1 (Proportional):** The hazard ratio remains constant across
  the entire timeline ($p > 0.05$).
- **Tier 2 (Time-Varying):** The impact of the threshold shifts
  significantly as time passes ($p < 0.05$). Here, $p = 0.000682$,
  meaning the hazard ratios fluctuate over time, suggesting the relative
  event risk evolves as follow-up increases.

Because we used the systematic grid engine on a 1-cut target, we can
plot the **1D optimization path curve** directly:

```{r plot-surface, fig.width=10, fig.height=5, fig.cap = "Systematic Log-Rank optimization curve. The peak denotes the precise coordinate location that maximizes separation."}
# Draws the line trajectory optimization peak
plot(cutpoint_result_adj, 
     type = "surface",
     title = "Custom Optimization Curve Title",
     xlab = "Biomarker Cut-point Threshold",
     ylab = "Chi-Square Test Statistic")
```

```{r find-cuts-plot, fig.cap = "Continuous population density profile with discovered optimal adjusted cut-point boundaries mapped as vertical anchors."}
plot(cutpoint_result_adj, type = "distribution")
```

------------------------------------------------------------------------

## 4. Bootstrap Stability Validation

To prove that our discovered threshold is a reproducible clinical signal
rather than an artifact of overfitting, we pass our adjusted model to
`validate_cutpoint()` for rigorous bootstrap resampling.

```{r validate-cuts}
validation_result_adj <- validate_cutpoint(
  # ==========================================
  # 1. VALIDATION INPUTS
  # ==========================================
  cutpoint_result = cutpoint_result_adj,
  num_replicates = 30, # Scaled for vignette performance; use >= 500 for publication
  n_cores = 1,
  # ==========================================
  # 2. AUTO-SCALING PASS-THROUGH TUNING
  # ==========================================
  seed = 123 # REPRODUCIBILITY
)

summary(validation_result_adj)
```

### 🔍 Automated Performance Grading: The 4-Tier Stability Assessment

Following bootstrap calculations, `OptSurvCutR` evaluates your
confidence interval spacing relative to your data distribution range:

```{r validate-cuts-plot, fig.cap = "Bootstrap discovery density profiles tracking threshold convergence across resampled patient pools."}
plot(validation_result_adj)
```

------------------------------------------------------------------------

### 4.1 Engineering Solutions: Deploying the `nmin` Wedge

Discrete clinical metrics (like positive lymph node integer
distributions) are rarely smooth, unskewed curves.

**The Overlap Problem:** In early baseline tuning passes with this raw
dataset, setting a low minimum sample threshold caused the algorithm to
group cut boundaries tightly around small count spaces. When subjected
to bootstrap resampling, the 95% confidence intervals spread
dramatically because discrete integer boundaries cause optimization
jumps. This high relative variance triggered an automated **Tier 4
(UNSTABLE)** flag.

**The Solution:** We can resolve this without changing our core clinical
objective by slightly increasing the minimum group size constraint
(`nmin`). Mandating that every individual cohort must hold a significant
proportion of the sample size stabilizes the internal Cox estimation
matrices, shrinking the bootstrap variance down cleanly.

------------------------------------------------------------------------

## 5. Visualisation and Interpretation

By default, the package maps the patient cohort into numeric groups
ordered sequentially based on your continuous predictor values:

\* **G1:** Low Lymph Node Burden

\* **G2:** Intermediate Lymph Node Burden

\* **G3:** High Lymph Node Burden

### Plot A: Covariate-Adjusted Kaplan-Meier Curves (`type = "outcome"`)

```{r plot-km, fig.width=8, fig.height=6, fig.cap = "Adjusted Kaplan-Meier overall survival estimates stratified by optimal positive node abundance cohorts."}
# 1. Inspect how many cut-points the Systematic Engine actually discovered
optimal_cuts <- sort(cutpoint_result_adj$optimal_cuts)
num_cuts <- length(optimal_cuts)

# 2. Dynamically assign matching label cohorts
if (num_cuts == 1) {
  dynamic_labels <- c("Low Nodal Burden", "High Nodal Burden")
} else if (num_cuts == 2) {
  dynamic_labels <- c("Low Burden", "Intermediate Burden", "High Burden")
} else {
  dynamic_labels <- paste("Cohort Group", 1:(num_cuts + 1))
}

# 3. Generate the adjusted Kaplan-Meier curve natively safely
plot(
  cutpoint_result_adj,
  type = "outcome",
  title = "5-Year Recurrence-Free Survival by Node Group",
  xlab = "Time (Months)",
  ylab = "Recurrence-Free Survival Probability",
  legend.title = "Nodal Burden Group",
  legend.labs = dynamic_labels
)
```

> **Interpretation:** The resulting survival paths reveal a clean
> degradation in recurrence-free survival matching your threshold.
> Patients in the **G1 (Low Burden)** tier track with significantly
> higher survival probability over the 60-month window compared to the
> high cluster ($58.3\%$ vs $29.8\%$). Because this grouping was
> uncovered using our covariate-adjusted framework, we can state with
> confidence that this survival divergence is an independent trait of
> the biomarker itself.

### Plot B: Adjusted Hazard Ratio Forest Plots (`type = "forest"`)

```{r plot-forest, fig.width=10, fig.height=3.5, fig.cap = "Multivariate Forest plot displaying adjusted biomarker Hazard Ratios alongside baseline clinical adjusters."}
plot(
  cutpoint_result_adj,
  type = "forest",
  reference_group = "G1",
  main = "Adjusted HRs for Lymph Node Groups (Ref: G1/Low)"
)
```

> **Interpretation:** This plot isolates the independent prognostic
> value required for clinical translation. Because `age`, `sex`, and
> `differ` are calculated simultaneously inside the model, the displayed
> Hazard Ratios represent pure, adjusted biological weights. High (G2)
> nodal presence is confirmed as an independent risk factor for
> recurrence mortality ($HR = 2.37$) relative to the low burden
> reference pool (G1).

------------------------------------------------------------------------

## 6. Advanced Diagnostics & Toolsets

### Plot C: Schoenfeld Residual Tracks (`type = "diagnostic"`)

```{r plot-residuals, fig.width=8, fig.height=5, fig.cap = "Schoenfeld residual diagnostic plot to verify the proportional hazards assumption."}
# The unexported S3 plotting engine automates the calculation of Schoenfeld residuals 
# by internally routing the optimized Cox model metrics into survival::cox.zph().
plot(cutpoint_result_adj, type = "diagnostic")
```

### The Package Data Prepper (`return_data = TRUE`)

If you wish to bypass standard package plotting routines and extract the
raw, cleaned data frame with your newly minted group assignments, set
`return_data = TRUE`.

```{r extract-data}
augmented_dataframe <- plot(cutpoint_result_adj, return_data = TRUE)

# View data mapping structure
head(augmented_dataframe[, c("factor", "time", "event", "group")])
```

------------------------------------------------------------------------

## 7. Conclusion & Next Steps

This vignette has demonstrated how to configure, discover, and validate
independent, covariate-adjusted survival thresholds using `OptSurvCutR`.
By factoring confounding adjusters into your grid search loops, you
ensure your downstream clinical risk stratification holds true
prognostic value.

- **Package Code Repository:**
  <https://github.com/paytonyau/OptSurvCutR>
- **Issue Tracker & Feature Requests:**
  <https://github.com/paytonyau/OptSurvCutR/issues>
- **Methodological Reference Manuscript:** Yau, Payton T. O.
  “OptSurvCutR: Validated Cut-point Selection for Survival Analysis.”
  *bioRxiv preprint*, 2025. <https://doi.org/10.1101/2025.10.08.681246>.

------------------------------------------------------------------------

## 8. Session Information

```{r session_info}
sessionInfo()
```

\`\`\`
