---
title: "Stratifying Colorectal Cancer Patients by Gut Enterovirus Abundance"
subtitle: "Survival Analysis Workflow with OptSurvCutR Demonstrating Covariate Adjustment"
author: "Payton Yau"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Stratifying Colorectal Cancer Patients by Gut Enterovirus Abundance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# Set global options for all code chunks in the document.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```


### About the Package

The **`OptSurvCutR`** package provides a comprehensive, data-driven workflow for survival analysis, specifically designed for scenarios where a single cut-point is not sufficient. Its core functionality is to discover and validate one or more optimal cut-points for continuous variables in time-to-event data. A key feature is the ability to perform these analyses while adjusting for covariates, allowing assessment of a biomarker's independent prognostic value.

This vignette focuses on demonstrating this covariate adjustment feature.

# Introduction

This tutorial demonstrates how to use the **`OptSurvCutR`** package to stratify patients into prognostic groups based on a continuous biomarker, specifically showing how to adjust for potential confounding variables.

### The Scientific Background


The human gut virome, the collection of viruses in the gut, can be disrupted (dysbiosis) in diseases like colorectal cancer (CRC). We will analyse gut virome data to determine if the relative abundance of the genus Enterovirus can predict 5-year overall survival, considering potential confounders like patient sex. This analysis, inspired by data from [Smyth et al. (2024)](https://doi.org/10.1002/cam4.70434), uses `OptSurvCutR` to identify data-driven thresholds for Enterovirus abundance both with and without adjustment for covariates.

### Analysis Workflow

This guide covers a complete and robust workflow:

1.  **Setup and Configuration**: Preparing the R environment and defining key parameters.
2.  **Data Loading and Preparation**: Loading the package's built-in datasets and preparing them for analysis.
3.  **Optimal Cut-point Discovery (Unadjusted vs. Adjusted)**: 
- Step 3.1: Determine the optimal number of groups (unadjusted and adjusted).
- Step 3.2: Find the cut-point locations (unadjusted and adjusted).
4.  **Cut-point Validation**: Assess the stability of the discovered thresholds via bootstrapping.
5.  **Visualisation and Interpretation**: Create and interpret Kaplan-Meier curves, forest plots, and other key visualisations.
6.  **Model Diagnostics**: Checking the assumptions of the survival model.

------------------------------------------------------------------------

## 1. Setup and Configuration

### Loading Libraries

First, we load all the necessary R packages.

```{r load-libraries}
# Load all necessary libraries for the analysis.
library(OptSurvCutR)
library(dplyr)
library(survival)
library(survminer) # For ggsurvplot, ggforest, ggcoxzph
library(ggplot2)
library(knitr) # For kable()
library(cli) # For enhanced messsexs
```

### Defining Parameters

We define the name of the microbe/virus we wish to analyse upfront. Here, we selected Enterovirus - a RNA virus within the Picornaviridae family, known for causing gastroenteritis and other infections. We included the covariates in the adjusted analysis.

```{r configuration}
# Define the biomarker of interest.
microbe_name_to_analyze <- "Enterovirus"

# Define covariates available in the crc_virome dataset
covariates_to_adjust_for <- c("sex")
```

------------------------------------------------------------------------

## 2. Data Loading and Preparation

We load the pre-cleaned clinical and virome dataset directly from the `OptSurvCutR` package. With the updated data format, this is now a single data frame. We will process it to extract the necessary columns, parse the survival status, and censor the data at 5 years (60 months) for this analysis.

```{r load-data}
# Load the single, wide-format data frame from the package.
data("crc_virome", package = "OptSurvCutR")

# --- Prepare the final analysis data frame ---
analysis_data <- crc_virome %>%
  # Select and rename key columns
  select(
    patient_id = sample_id,
    time_months,
    status_char = status, # Original character status (e.g., "0:LIVING")
    abundance = all_of(microbe_name_to_analyze),
    all_of(covariates_to_adjust_for) # Include covariates
  ) %>%
  # Apply 5-year (60-month) censoring and parse status
  mutate(
    # 1. Parse numeric status (0/1) from the character string
    status_numeric = as.numeric(substr(status_char, 1, 1)),

    # 2. Apply censoring: events after 60 months are set to 'censored' (0)
    status_final = ifelse(time_months > 60, 0, status_numeric),

    # 3. Cap all follow-up times at 60 months
    time_final = pmin(time_months, 60)
  ) %>%
  # Select final columns for analysis
  select(
    patient_id,
    time = time_final,
    status = status_final, # Use the new 5-year censored status
    abundance,
    all_of(covariates_to_adjust_for)
  ) %>%
  # Remove rows with any missing values in essential columns
  filter(complete.cases(time, status, abundance, across(all_of(covariates_to_adjust_for))))

# Display the head of the final processed data
head(analysis_data)
```

------------------------------------------------------------------------

## 3. Optimal Cut-point Discovery (Unadjusted vs. Adjusted)

**A Note on Reproducibility (seed)**: Functions involving the genetic algorithm (`method = "genetic"`) or bootstrapping (`validate_cutpoint`) use random processes. We set the seed argument to ensure results are reproducible.

### Step 3.1: Determine the Optimal Number of Cut-points

We first run `find_cutpoint_number()` without covariates, then with covariates, to see if adjustment changes the suggested number of groups. We use BIC and the genetic algorithm (checking if `rgenoud` is available).

```{r find-number-Unadjusted-Analysis}
# --- Unadjusted Analysis ---
# --- Unadjusted Analysis ---
number_result_unadj <- find_cutpoint_number(
  data = analysis_data,
  predictor = "abundance",
  outcome_time = "time",
  outcome_event = "status",
  method = "genetic",
  criterion = "BIC",
  max_cuts = 4, # Test 0-3 cuts (4 groups max)
  nmin = 0.1, # Min 10% per group
  max.generations = 50, # Reduced for vignette speed
  pop.size = 50, # Reduced for vignette speed
  seed = 42
)
```

```{r find-number-Adjusted-Analysis}
# --- Adjusted Analysis ---
number_result_adj <- find_cutpoint_number(
  data = analysis_data,
  predictor = "abundance",
  outcome_time = "time",
  outcome_event = "status",
  method = "genetic",
  criterion = "BIC",
  max_cuts = 4,
  nmin = 0.1,
  max.generations = 50,
  pop.size = 50,
  covariates = covariates_to_adjust_for, # ADDED COVARIATES
  seed = 43 # Use a different seed
)
```

```{r find-number-plots, fig.alt = "Two line plots comparing Bayesian Information Criterion (BIC) against the number of cut-points. The left plot shows unadjusted analysis, and the right shows analysis adjusted for sex. Both show a minimum at 2 cut-points."}
plot_num_unadj <- plot(number_result_unadj) +
  ggtitle("BIC vs Cuts (Unadj.)") +
  theme(plot.title = element_text(size = 14))

plot_num_adj <- plot(number_result_adj) +
  ggtitle(paste0(
    "BIC vs Cuts (Adj. for ",
    paste(covariates_to_adjust_for, collapse = ", "), ")"
  )) +
  theme(plot.title = element_text(size = 14))

# Show plots sequentially
print(plot_num_unadj)
print(plot_num_adj)
```
> **Interpretation:** We compare the BIC plots and optimal number of cuts from both the unadjusted and adjusted analyses. In this example, both might suggest 2 cut-points (3 groups) remain optimal even after adjusting for sex. If the optimal number changed significantly after adjustment, it might suggest the biomarker's grouping effect is partly confounded. We proceed using the optimal number suggested by the adjusted analysis.

### Step 3.2: Find the Location of the Cut-points

Now that we have decided to use two cut-points, we use `find_cutpoint()` to identify their optimal values based on the log-rank statistic.

```{r find-cuts-Unadjusted-Analysis}
# Determine optimal number for subsequent steps based on unadjusted result
optimal_n_cuts_unadj <- number_result_unadj$optimal_num_cuts # Store unadjusted for comparison

# --- Unadjusted Analysis ---
cutpoint_result_unadj <- find_cutpoint(
  data = analysis_data,
  predictor = "abundance",
  outcome_time = "time",
  outcome_event = "status",
  method = "genetic",
  criterion = "logrank",
  num_cuts = optimal_n_cuts_unadj,
  nmin = 0.1,
  max.generations = 50,
  pop.size = 50,
  seed = 123
)

print(summary(cutpoint_result_unadj))

optimal_cuts_unadj_values <- cutpoint_result_unadj$optimal_cuts
```

```{r find-cuts-Adjusted-Analysis}
# Determine optimal number for subsequent steps based on ADJUSTED result
optimal_n_cuts_adj <- number_result_adj$optimal_num_cuts

# --- Adjusted Analysis ---
cutpoint_result_adj <- find_cutpoint(
  data = analysis_data,
  predictor = "abundance",
  outcome_time = "time",
  outcome_event = "status",
  method = "genetic",
  criterion = "logrank",
  num_cuts = optimal_n_cuts_adj,
  nmin = 0.1,
  max.generations = 50,
  pop.size = 50,
  covariates = covariates_to_adjust_for,
  seed = 124 # Different seed
)

print(summary(cutpoint_result_adj))

optimal_cuts_adj_values <- cutpoint_result_adj$optimal_cuts
```

```{r find-cuts-plot, fig.alt = "Two density plots of Enterovirus abundance. Vertical lines indicate the optimal cut-points. The left plot shows unadjusted cut-points, while the right shows cut-points adjusted for sex."}
# Plot comparison
plot_dist_unadj <- plot(cutpoint_result_unadj,
  type = "distribution"
) +
  ggtitle("Unadj. Cut-points")

plot_dist_adj <- plot(cutpoint_result_adj,
  type = "distribution"
) +
  ggtitle(paste(
    "Adj. Cut-points (for",
    paste(covariates_to_adjust_for, collapse = ", "), ")"
  ))

# Show plots sequentially
print(plot_dist_unadj)
print(plot_dist_adj)
```

> **Interpretation**: We compare the optimal cut-point locations found with and without covariate adjustment. If the cut-points shift substantially after adjustment, it suggests the covariates influenced the optimal thresholds. If they remain similar, it strengthens the evidence for the biomarker's independent grouping effect. We will use the adjusted cut-points for the subsequent validation and final modeling.

------------------------------------------------------------------------

## 4. Cut-point Validation

This is a crucial step to assess how reliable our discovered cut-points are. We use `validate_cutpoint()` to run a **bootstrap analysis**. This re-runs the `find_cutpoint()` algorithm many times on resampled data to see how much the optimal cut-points vary.

```{r validate-cuts}
validation_result_adj <- validate_cutpoint(
  cutpoint_result = cutpoint_result_adj, # Use ADJUSTED result
  num_replicates = 25, # REDUCED for vignette speed; use >= 500 for real analyses
  n_cores = 1,
  max.generations = 50, # Passed via ... to find_cutpoint
  pop.size = 50, # Passed via ... to find_cutpoint
  seed = 456
  # Note: validate_cutpoint implicitly uses the covariates from cutpoint_result_adj
)
```

```{r validate-cuts-plot, fig.alt = "Density plot showing the distribution of cut-points derived from bootstrap replicates. The peaks align closely with the original cut-points (solid lines), indicating stability."}
bootstrap_plot_adj <- plot(validation_result_adj) +
  ggtitle("Bootstrap Stability of Adjusted Cut-points")
print(bootstrap_plot_adj)
```
> **Interpretation:** The bootstrap results assess the stability of the adjusted cut-points. Narrow confidence intervals indicate that even after accounting for sex, the thresholds defining the Low, Medium, and High Enterovirus abundance groups are reasonably robust to sampling variation.

------------------------------------------------------------------------

## 5. Visualisation and Interpretation (Using Adjusted Results)

We create the final prognostic groups based on the adjusted optimal cuts.

```{r categorize-data}
# Create a new column with the adjusted abundance group
analysis_data_categorised <- analysis_data %>%
  mutate(abundance_group_adj = cut(
    abundance,
    breaks = c(-Inf, optimal_cuts_adj_values, Inf),
    labels = c("Low", "Medium", "High"),
    right = FALSE
  )) %>%
  mutate(abundance_group_adj = factor(abundance_group_adj,
    levels = c("Low", "Medium", "High")
  ))

# Set "Medium" abundance as the reference group for the Cox model
analysis_data_categorised$abundance_group_adj <- relevel(
  analysis_data_categorised$abundance_group_adj,
  ref = "Medium"
)
```

### Fit Final Adjusted Models
We fit the final Kaplan-Meier and Cox models, including the covariates in the Cox model to get adjusted hazard ratios.

```{r plot-km2}
# Fit Kaplan-Meier model (usually plotted without covariate adjustment)
km_fit_adj <- survfit(Surv(time, status) ~ abundance_group_adj,
  data = analysis_data_categorised
)

# Fit the final ADJUSTED Cox Proportional-Hazards model
covariate_formula_part <- paste(covariates_to_adjust_for, collapse = " + ")
# Create the formula as a string
formula_string <- paste(
  "Surv(time, status) ~ abundance_group_adj +",
  covariate_formula_part
)
# Pass the string to as.formula() *inside* the coxph call
cox_model_adj <- coxph(as.formula(formula_string),
  data = analysis_data_categorised
)
```

### Plot A: Kaplan-Meier Curve

A Kaplan-Meier curve shows the probability of survival over time for different patient groups.

```{r plot-km, fig.width=8, fig.height=6, fig.alt = "Kaplan-Meier survival curves for three groups: Low, Medium, and High Enterovirus abundance. The Medium group shows the highest survival probability over 60 months, while Low and High groups show lower survival."}
km_plot_adj <- ggsurvplot(
  km_fit_adj,
  data = analysis_data_categorised,
  pval = TRUE,
  risk.table = TRUE,
  legend.title = paste(microbe_name_to_analyze, "(Adj. Groups)"),
  legend.labs = c("Medium", "Low", "High"), # Assumes Medium is ref level
  palette = c("#2E9FDF", "#E7B800", "#FC4E07"), # Blue, Yellow, Red
  ylim = c(0.4, 1.0),
  pval.coord = c(0, 0.45),
  title = "5-Year OS by Adjusted Enterovirus Group",
  xlab = "Time (Months)",
  ylab = "Overall Survival Probability",
  risk.table.title = "Number at risk"
)

print(km_plot_adj)
```

> **Interpretation:** The Kaplan-Meier plot using groups defined by the adjusted cut-points still shows a clear separation, confirming the U-shaped risk pattern persists even after considering sex. The Medium group maintains the best prognosis, while both Low and High groups have significantly worse survival (log-rank p < 0.05). Comparing this visually to an unadjusted KM plot (not shown here, but could be generated using optimal_cuts_unadj_values) can reveal subtle shifts caused by covariate adjustment.

### Plot B: Forest Plot of Hazard Ratios

This forest plot shows the Hazard Ratios (HR) for the Low and High abundance groups relative to the Medium group, adjusted for sex.

```{r plot-forest, fig.width=8.5, fig.height=4.5, fig.alt = "Forest plot displaying Hazard Ratios. Compared to the Medium group (reference), both Low and High abundance groups have Hazard Ratios significantly greater than 1."}
forest_plot_adj <- ggforest(
  cox_model_adj,
  data = analysis_data_categorised,
  main = paste(
    "Adjusted HRs for",
    microbe_name_to_analyze,
    "Groups (Ref: Medium)"
  )
)

print(forest_plot_adj)
```

> **Interpretation:** This plot quantifies the independent prognostic value of the Enterovirus abundance groups after controlling for sex.
> - The Hazard Ratios for the Low and High groups are still significantly greater than 1.0 (confidence intervals do not cross 1), indicating that both low and high abundance remain associated with increased mortality risk independently of sex.
> - Compare these adjusted HRs to the unadjusted HRs (from the previous summary(cutpoint_result_unadj) or a separate unadjusted ggforest plot). If the adjusted HRs are closer to 1.0 than the unadjusted ones, it suggests sex partially confounded the original association. If they remain similar and significant, it strengthens the evidence for the biomarker's independent effect.
> - The plot also shows the adjusted HRs for sex themselves.

------------------------------------------------------------------------

## 6. Model Diagnostics

### Plot D: Biomarker Distribution by Risk Group

This box plot confirms how the `find_cutpoint()` function has partitioned the patients based on their abundance levels.

```{r plot-boxplot, fig.width=8, fig.height=5, fig.alt = "Boxplots showing the distribution of Enterovirus abundance within the three identified risk groups (Low, Medium, High)."}
dist_plot_adj <- ggplot(
  analysis_data_categorised,
  aes(
    x = abundance_group_adj,
    y = abundance,
    fill = abundance_group_adj
  )
) +
  geom_boxplot(
    show.legend = FALSE,
    outlier.shape = NA
  ) +
  geom_jitter(
    width = 0.2,
    alpha = 0.3,
    size = 1,
    show.legend = FALSE
  ) +
  labs(
    title = paste(
      "Distribution of",
      microbe_name_to_analyze,
      "Abundance by Adjusted Risk Group"
    ),
    x = "Adjusted Prognostic Group",
    y = "Relative Abundance" # Check if log scale needed
  ) +
  scale_fill_manual(values = c("#E7B800", "#2E9FDF", "#FC4E07")) +
  theme_minimal(base_size = 12) +
  coord_cartesian(ylim = quantile(analysis_data_categorised$abundance,
    c(0.01, 0.99),
    na.rm = TRUE
  )) # Zoom axis

print(dist_plot_adj)
```

### Plot E: Checking the Proportional Hazards Assumption

A key assumption of the Cox model is that the hazard ratio (the effect of the predictor) is constant over time. We check this using Schoenfeld residuals. A non-significant p-value (p > 0.05) for the GLOBAL test supports this assumption.

```{r plot-residuals, fig.width=8, fig.height=5, fig.alt = "Plots of Schoenfeld residuals over time for the Cox model variables. The smoothed trend lines are approximately horizontal, supporting the proportional hazards assumption."}
schoenfeld_test_adj <- cox.zph(cox_model_adj)

print(schoenfeld_test_adj)
print(ggcoxzph(schoenfeld_test_adj))
```

> **Interpretation**: The `cox.zph` test provides a formal statistical check. If the p-value for `GLOBAL` is greater than 0.05, we generally conclude that there is no strong evidence against the proportional hazards assumption. The plot provides a visual check; ideally, the smoothed lines should be roughly horizontal with confidence bands overlapping zero. Minor deviations are often acceptable, especially with smaller group sizes.

------------------------------------------------------------------------

## 7. Conclusion & Next Steps

This vignette has demonstrated the three-step workflow for cut-point analysis using `OptSurvCutR`. By following this workflow, users can confidently identify and validate robust, statistically-optimal thresholds in their own survival data, moving beyond simple median splits to uncover more nuanced relationships.

We encoursex you to try `OptSurvCutR` with your own data.

-   **Install the package** from GitHub: `remotes::install_github("paytonyau/OptSurvCutR")`
-   **Report issues or suggest features** on our [GitHub psex](https://github.com/paytonyau/OptSurvCutR).
-   **Star the repository** if you find it useful.
- Cite the package: Please cite the accompanying paper if you use OptSurvCutR in your research: Yau, Payton T. O. “OptSurvCutR: Validated Cut-point Selection for Survival Analysis.” bioRxiv preprint, posted October 18, 2025. https://doi.org/10.1101/2025.10.08.681246.
- If you find OptSurvCutR useful for your research, please consider supporting its ongoing development and maintenance. Your contribution helps keep the project alive and improving!
```

------------------------------------------------------------------------

## 8. Session Information

For reproducibility, the session information below lists the R version and all attached packages.

```{r session_info}
sessionInfo()
```
