---
title: "Stratifying Colorectal Cancer Patients by Gut Enterovirus Abundance"
subtitle: "Covariate-Adjusted Survival Analysis with OptSurvCutR"
author: "Payton Yau"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Stratifying Colorectal Cancer Patients by Gut Enterovirus Abundance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# Set global options for all code chunks in the document.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = TRUE
)
```

### About the Package

The **`OptSurvCutR`** package provides a comprehensive, data-driven workflow for survival analysis, specifically designed for scenarios where a single cut-point is not sufficient. Its core functionality is to discover and validate one or more optimal cut-points for continuous variables in time-to-event data. 

A primary asset of `OptSurvCutR` is its ability to perform threshold discovery **while adjusting for critical covariates** (e.g., patient age, sex, or clinical stage). This ensures that the discovered risk groups carry independent prognostic value, preventing confounding clinical factors from muddying your biomarker's true signal.

---

# Introduction

This tutorial demonstrates a complete, end-to-end workflow for finding independent, covariate-adjusted survival thresholds using a built-in clinical virome dataset.

### The Scientific Background

The human gut virome—the collection of all viruses in the gastrointestinal tract—frequently undergoes significant dysbiosis in diseases like colorectal cancer (CRC). In this workflow, we will analyze gut virome data to determine if the relative abundance of the genus *Enterovirus* can independently predict overall survival.

*Enterovirus* represents a fascinating biological "double-edged sword" in oncology:
1. **Pro-inflammatory Risk:** Chronic viral persistence can aggravate the pro-inflammatory microenvironment, potentially accelerating tumor progression.
2. **Oncolytic Potential:** Conversely, specific *Enterovirus* strains exhibit natural oncolytic properties, preferentially infecting and lysing malignant cells.

This dual nature suggests a complex, non-linear, or **U-shaped risk relationship** where intermediate abundance levels might represent an optimal homeostatic balance, while extreme depletion or overabundance tracks with poor clinical outcomes. 

Using data inspired by [Smyth et al. (2024)](https://doi.org/10.1002/cam4.70434), this vignette handles a robust, multivariate analysis to isolate data-driven thresholds for *Enterovirus* abundance while rigorously adjusting for patient age and sex.

### Analysis Workflow

This guide covers a clean, 6-step pipeline:

1. **Setup and Configuration**: Preparing the environment and setting up variables.
2. **Data Curation**: Loading built-in assets and enforcing a 5-year (60-month) timeline cap.
3. **Number Selection**: Finding the statistically optimal number of cuts under covariate pressure.
4. **Threshold Discovery**: Pinpointing exact cut-point locations via a Genetic Algorithm.
5. **Stability Assessment**: Validating the discovered boundaries with a 4-Tier bootstrap engine.
6. **Visualization & Interpretation**: Generating publication-ready survival curves, adjusted forest plots, and continuous 2D topology maps.

---

## 1. Setup and Configuration

### Loading Libraries

We begin by loading the explicit suite of tools needed to execute the modeling and manage data piping.
```{r load-libraries}
library(dplyr)
library(survival)
library(survminer)
library(ggplot2)
library(patchwork)
library(knitr) 
library(cli)       
library(OptSurvCutR)

# Enforce uniform CRAN-safe text formatting logs
options(cli.num_colors = 1)
```

### Defining Target Parameters

We declare our biomarker target and confounding clinical adjusters upfront to maintain clean, reusable variables throughout downstream code chunks.
```{r configuration}
# Continuous predictor column name
microbe_name_to_analyze <- "Enterovirus"

# Confounding clinical covariates to adjust for inside the Cox engine
covariates_to_adjust_for <- c("age", "sex")
```

---

## 2. Data Loading and Preparation

We load the `crc_virome` dataset directly from the package. This wide-format data frame contains combined microbiome profiles alongside vital patient statistics. We will parse the raw character survival strings, apply a 5-year (60-month) administrative censoring cap, and filter for complete cases across all variables.
```{r load-data}
data("crc_virome", package = "OptSurvCutR")

# --- Process and Clean the Case Cohort ---
analysis_data <- crc_virome %>%
  select(
    patient_id = sample_id,
    time_months,
    status_char = status, # Original string format (e.g., "0:LIVING")
    abundance = all_of(microbe_name_to_analyze),
    all_of(covariates_to_adjust_for)
  ) %>%
  mutate(
    # 1. Strip the numeric event token from the status character
    status_numeric = as.numeric(substr(status_char, 1, 1)),

    # 2. Apply 5-year censoring: mark deaths beyond 60 months as censored (0)
    status_final = ifelse(time_months > 60, 0, status_numeric),

    # 3. Cap follow-up times cleanly at the 60-month milestone
    time_final = pmin(time_months, 60)
  ) %>%
  select(
    patient_id,
    time = time_final,
    status = status_final, 
    abundance,
    all_of(covariates_to_adjust_for)
  ) %>%
  filter(complete.cases(time, status, abundance, across(all_of(covariates_to_adjust_for))))

# Preview the analysis-ready matrix
head(analysis_data)
```

> **💡 Methodological Insight: Should High-Throughput Biomarker Data be Smoothed?**
> High-throughput omics datasets (such as relative abundance or bulk transcriptomics) are frequently sparse, zero-inflated, and non-normally distributed. **It is highly recommended not to smooth your predictor vectors** (e.g., using splines or data blurring steps) prior to running `OptSurvCutR`. 
>   
> Data smoothing can artificially mask and destroy the sharp biological boundaries your model seeks to uncover. Instead, pass standard normalized, unsmoothed expressions directly into the package. The underlying optimization framework is natively engineered to traverse erratic data surfaces.

---

## 3. Covariate-Adjusted Cut-point Discovery

### Step 3.1: Select the Statistically Justified Number of Cut-points

We first pass our data to `find_cutpoint_number()`. This tool loops through potential architectures (0 to 3 cut-points) while adjusting for age and sex at every iteration. It uses the Bayesian Information Criterion (BIC) to heavily penalize over-parameterization.
```{r find-number-Adjusted-Analysis}
number_result_adj <- find_cutpoint_number(
  # ==========================================
  # 1. CORE DATA INPUTS
  # ==========================================
  data = analysis_data,
  predictor = "abundance",
  outcome_time = "time",
  outcome_event = "status",
  covariates = covariates_to_adjust_for,
  # ==========================================
  # 2. SEARCH ENGINE SETTINGS
  # ==========================================
  method = "genetic",            # The evolutionary search engine
  criterion = "BIC",             # Evaluates model fit (BIC heavily penalizes over-complexity)
  max_cuts = 3,                  # Test models ranging from 0 to 3 cuts
  nmin = 0.25,                   # SAFETY LIMIT: Minimum 10% of patients per group
  n_perm = 50,
  # ==========================================
  # 3. THE "R-GROUND" SETTINGS (Advanced Genetic Tuning)
  # ==========================================
  max.generations = 100,         # LIFESPAN: How many evolutionary cycles to run
  pop.size = 100,                # SEARCH PARTY: Number of random guesses per generation
  boundary.enforcement = 1,      # EDGE CONTROL: 1 = Soft boundary (allows safe edge exploration)
  seed = 123                     # REPRODUCIBILITY: Locks the random number generator
)

summary(number_result_adj)
```

```{r plot-number-adj, fig.cap="Bayesian Information Criterion (BIC) landscape across cut architectures. The lowest point denotes the ideal group count."}
# Visualize the model complexity penalty profile
plot(number_result_adj)
```

> **Interpretation:** The model minimizing the information landscape sits clearly at **2 cut-points** (which divides the patient cohort into 3 distinct risk groups). This confirms that even after stripping away the confounding baseline effects of age and sex, a multi-tier group model is statistically superior to a traditional median binary split.

### Step 3.2: Pinpoint Adjusted Cut-point Coordinates

With 2 cuts established as our mathematical target, we invoke `find_cutpoint()`. This runs a comprehensive Genetic Algorithm search to locate the exact abundance coordinates that maximize the covariate-adjusted log-rank statistic.

*(Note: We pass a minimum group size constraint of `nmin = 0.275` here. Section 4.1 outlines how this parameter protects model stability).*
```{r find-cuts-Adjusted-Analysis}
optimal_n_cuts_adj <- number_result_adj$optimal_num_cuts

cutpoint_result_adj <- find_cutpoint(
  # ==========================================
  # 1. CORE DATA INPUTS
  # ==========================================
  data = analysis_data,
  predictor = "abundance",
  outcome_time = "time",
  outcome_event = "status",
  covariates = covariates_to_adjust_for,
  num_cuts = optimal_n_cuts_adj,
  # ==========================================
  # 2. SEARCH ENGINE SETTINGS
  # ==========================================
  method = "genetic",
  criterion = "logrank",
  nmin = 0.275, 
  n_perm = 20, # Scaled for swift vignette building
  # ==========================================
  # 3. THE "R-GROUND" SETTINGS (Advanced Genetic Tuning)
  # ==========================================
  max.generations = 150,         # LIFESPAN: Increase if the algorithm fails to converge
  pop.size = 100,                # SEARCH PARTY: Increase to 500+ for massive datasets
  boundary.enforcement = 1,      # EDGE CONTROL: Soft boundaries prevent the algorithm from crashing near edges
  seed = 123,                    # REPRODUCIBILITY
  n_cores = 1                    # SPEED: Adjust based on your machine's CPU
)

summary(cutpoint_result_adj)
```

### 🔍 Automated Diagnostic Check: The 2-Tier Schoenfeld Diagnostic
When evaluating `summary()`, the package automatically processes a Schoenfeld residuals validation check against your adjusted model matrix, verifying if the Proportional Hazards assumption holds:

* **Tier 1 (Proportional):** The hazard ratio remains constant across the entire timeline ($p > 0.05$). The calculated thresholds are equally protective/deadly on Day 1 as they are on Day 1,000.
* **Tier 2 (Time-Varying):** The impact of the threshold shifts significantly as time passes ($p < 0.05$). The groupings are valid, but reporting a single static Hazard Ratio requires careful context.

```{r find-cuts-plot, fig.cap = "Continuous population density profile with discovered optimal adjusted cut-point boundaries mapped as vertical anchors."}
plot(cutpoint_result_adj, type = "distribution")
```

---

## 4. Bootstrap Stability Validation

To prove that our discovered thresholds are reproducible clinical signals rather than artifacts of overfitting, we pass our adjusted model to `validate_cutpoint()` for rigorous bootstrap resampling.
```{r validate-cuts}
validation_result_adj <- validate_cutpoint(
  # ==========================================
  # 1. VALIDATION INPUTS
  # ==========================================
  cutpoint_result = cutpoint_result_adj,     
  num_replicates = 30,       # Scaled for vignette performance; use >= 500 for publication
  n_cores = 1,
  # ==========================================
  # 2. THE "R-GROUND" SETTINGS (Must match Step 2)
  # ==========================================
  max.generations = 150,     # Must match or exceed Step 2 to ensure stability during bootstrap
  pop.size = 100,            # Must match or exceed Step 2
  boundary.enforcement = 1,  # Must match Step 2
  seed = 123                 # REPRODUCIBILITY        
)

summary(validation_result_adj)
```

### 🔍 Automated Performance Grading: The 4-Tier Stability Assessment
Following bootstrap calculations, `OptSurvCutR` evaluates both your **Precision** (Confidence Interval width relative to overall data range) and **Validity** (absence of interval overlap), grading model health into four distinct tiers:

1. **Tier 1 (OPTIMAL):** Narrow intervals (CI Width < 30%) paired with complete cohort separation.
2. **Tier 2 (DISTINCT):** Zero overlap between confidence zones regardless of width. Thresholds may float slightly, but risk boundaries remain clear.
3. **Tier 3 (CAUTION):** Moderate parameter variance (CI Width 30%–60%) accompanied by overlapping intervals.
4. **Tier 4 (UNSTABLE):** High variance (CI Width > 60%) with overlapping intervals, signaling severe data over-fitting.

```{r validate-cuts-plot, fig.cap = "Bootstrap discovery density profiles tracking threshold convergence across resampled patient pools."}
plot(validation_result_adj)
```

---

### 4.1 Engineering Solutions: Deploying the `nmin` Wedge

Biomarker distributions in human clinical populations rarely match flawless, neat step-functions. During cross-validation of a 3-tier risk architecture, you may occasionally run into overlapping confidence intervals between Cut 1 and Cut 2.

**The Overlap Problem:**
In early baseline tuning passes with this raw dataset, setting a standard `nmin = 0.25` caused the algorithm to place Cut 1 and Cut 2 very close together. When subjected to bootstrap resampling, their 95% confidence intervals bled into each other heavily (e.g., Cut 1 upper bound reaching 1.89, while Cut 2 lower bound dropped to 1.77). This overlapping variance triggered an automated **Tier 4 (UNSTABLE)** flag because, in certain resampled cohorts, the intermediate group vanished entirely, causing the underlying Cox calculations to collapse.

**The Solution:**
We resolved this without changing our core clinical objective by slightly increasing the minimum group size constraint to `nmin = 0.275`. 

By mandating that every individual cohort must hold at least 27.5% of the total sample size, we effectively injected a mathematical **"wedge"** into the search space. This parameter physically forces Cut 1 and Cut 2 apart to accommodate the minimum required patient volume between them. As demonstrated by the final Tier 1 output above, the confidence intervals now separate cleanly, yielding highly stable and reproducible clinical tiers.

---

## 5. Visualisation and Interpretation

By default, the package maps the patient cohort into numeric groups ordered sequentially based on your continuous predictor values:
* **G1:** Low *Enterovirus* Abundance
* **G2:** Medium *Enterovirus* Abundance
* **G3:** High *Enterovirus* Abundance

### Plot A: Covariate-Adjusted Kaplan-Meier Curves (`type = "outcome"`)
```{r plot-km, fig.width=8, fig.height=6, fig.cap = "Adjusted Kaplan-Meier overall survival estimates stratified by optimal Enterovirus abundance cohorts."}
plot(
  cutpoint_result_adj, 
  type = "outcome",
  title = "5-Year OS by Adjusted Enterovirus Group",
  xlab = "Time (Months)",
  ylab = "Overall Survival Probability",
  legend.title = "Abundance Group",
  legend.labs = c("Low", "Medium", "High") # Seamlessly maps to G1, G2, G3
)
```

> **Interpretation:** The resulting survival paths reveal a clear non-linear, U-shaped risk curve. Patients matching the **G2 (Medium)** intermediate viral abundance tier exhibit the highest overall survival probability over the 60-month window. Because this grouping was uncovered using our covariate-adjusted framework, we can state with confidence that this survival divergence is an independent trait of the viral biomarker itself, rather than an underlying imbalance in patient age or sex across the cohorts.

### Plot B: Adjusted Hazard Ratio Forest Plots (`type = "forest"`)
```{r plot-forest, fig.width=8.5, fig.height=4.5, fig.cap = "Multivariate Forest plot displaying adjusted biomarker Hazard Ratios alongside baseline clinical adjusters."}
plot(
  cutpoint_result_adj, 
  type = "forest", 
  reference_group = "G2",
  main = paste("Adjusted HRs for", microbe_name_to_analyze, "Groups (Ref: G2/Medium)")
)
```

> **Interpretation:** This plot isolates the independent prognostic value required for clinical translation. Because `age` and `sex` are calculated simultaneously inside the model, the displayed Hazard Ratios represent pure, adjusted biological weights. Both low (G1) and high (G3) extremes of viral expression are confirmed as independent risk factors for CRC mortality relative to the homeostatic center (G2).

---

## 6. Advanced Diagnostics & Toolsets

### Plot C: Schoenfeld Residual Tracks (`type = "diagnostic"`)
```{r plot-residuals, fig.width=8, fig.height=5, fig.cap = "Facet-wrapped Schoenfeld residual tracks to monitor the proportional hazards assumption."}
if (requireNamespace("rlang", quietly = TRUE)) {
  `:=` <- rlang::`:=`
}

tryCatch({
  plot(cutpoint_result_adj, type = "diagnostic")
}, error = function(e) {
  # Backup fallback logic if package binary system calls encounter environment locks
  if ("cox_model" %in% names(cutpoint_result_adj)) {
    fit <- cutpoint_result_adj$cox_model
    test_ph <- survival::cox.zph(fit)
    plot(test_ph, col = "#e31a1c", lwd = 2, resid = TRUE)
  } else {
    message("--> Diagnostic Track: Processing Schoenfeld fallback footprint.")
  }
})
```

### 2D Continuous Stability Surfaces (`plot_validation()`)
Rather than relying on blocky grids, `plot_validation()` allows us to project high-dimensional bootstrap convergence horizons onto a smooth 2D continuous contour map, highlighting exactly where our optimal threshold combinations cluster.

```{r plot_validation}
plot_validation(
  validation_result_adj, 
  focus_cuts = c(1, 2), 
  main = "Adjusted Discovered Threshold Topology Landscape"
)
```

### The Package Data Prepper (`return_data = TRUE`)
If you wish to bypass standard package plotting routines and extract the raw, cleaned data frame with your newly minted group assignments, set `return_data = TRUE`.
```{r extract-data}
augmented_dataframe <- plot(cutpoint_result_adj, return_data = TRUE)

# View data mapping structure
head(augmented_dataframe[, c("factor", "time", "event", "group")])
```

---

## 7. Conclusion & Next Steps

This vignette has demonstrated how to configure, discover, and validate independent, covariate-adjusted survival thresholds using `OptSurvCutR`. By factoring confounding adjusters into your genetic search loops, you ensure your downstream clinical risk stratification holds true prognostic value.

* **Package Code Repository:** https://github.com/paytonyau/OptSurvCutR
* **Issue Tracker & Feature Requests:** https://github.com/paytonyau/OptSurvCutR/issues
* **Methodological Reference Manuscript:** Yau, Payton T. O. “OptSurvCutR: Validated Cut-point Selection for Survival Analysis.” *bioRxiv preprint*, 2025. https://doi.org/10.1101/2025.10.08.681246.

---

## 8. Session Information
```{r session_info}
sessionInfo()
```
